Comparison of several sequential and parallel derivative-free global optimization algorithms

### 2018-11-28 2019/PAVT/pavt_paper.tex
PAVT: add figures

\documentclass{svproc}
%
% RECOMMENDED
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%
%
\usepackage{graphicx}
\usepackage{marvosym}
\usepackage{amssymb}
\usepackage{cite}
% to typeset URLs, URIs, and DOIs
\usepackage{url}
\usepackage{hyperref}
\def\UrlFont{\rmfamily}

@@

\begin{abstract}
This work considers several stochastic and deterministic derivative-free global optimization
algorithms. In the first part of the paper popular sequential open-source solvers are compared against
the Globalizer solver, which is developed at the Lobachevskii State University. The Globalizer is
designed to solve problems with black-box objectives satisfying the Lipschitz condition and shows
competitive performance with other similar solvers. The comparison is done on several sets of
challenging multi-extremal benchmark functions. The second part of this work is devoted to
comparison between the Globalizer and MIDACO solvers on systems with shared and distributed
memory. MIDACO is a state-of-the-art global solver included to the TOMLAB optimization
environment for MATLAB. Results of the benchmark show advantages of the Globalizer on small-
dimensional, but sufficiently multi-extremal benchmark functions.
% We would like to encourage you to list your keywords within
% the abstract section using the \keywords{...} command.
\keywords{deterministic global optimization $\cdot$ stochastic global optimization
  $\cdot$ parallel numerical methods $\cdot$ derivative-free algorithms $\cdot$ black-box
optimization}
\end{abstract}
%
\section{Introduction}

@@

Let us consider the averaged number of trials executed to solve a single problem and the number of
solved problems as the characteristics of the optimization method on each class. The less the number
of trials, the faster the method converges to a solution, hence the less times it turns to a potentially
computation-costly procedure of computing the objective function. The number of solved problems
evidences the reliability of the method at given parameters on the class of test problems being
solved. In order to calculate the quantities featuring the reliability and the speed of convergence,
when calculating the averaged number of trials the solved problems were taken into account only.

@@

The computational experiments with parallel algorithms have been carried out on the Lobachevskii
supercomputer at
State University of Nizhni Novgorod. A computational node includes 2 Intel
Sandy Bridge E5-2660 2.2 GHz processors, 64 GB RAM. Each CPU has 8 cores (i. e. total 16
cores were available per a node). All computational nodes are working under CentOS Linux 7.2.
All considered algorithms are implemented using C++ and compiled by GCC 4.8.5.

@@

\begin{table}
  \centering
  \caption{Averaged numbers of iterations executed by the parallel algorithm for solving the test
optimization problems}
  \label{tab:iterations}
  \begin{tabular}{cccccccc}
    \cline{3-8}\noalign{\smallskip}
    \multicolumn{2}{c}{  } & \textit{p} & \multicolumn{2}{c}{Globalizer} & &
\multicolumn{2}{c}{MIDACO}   \\
    \noalign{\smallskip} \cline{4-5} \cline{7-8}  \noalign{\smallskip}
    \multicolumn{2}{c}{  } & & \textit{Simple} & \textit{Hard} & & \textit{Simple} &
\textit{Hard}  \\
    \noalign{\smallskip}\hline
    I & \textbf{1 cluster node}
      & \textit{1} &   25270 (100)  & 55180 (99) & & 27645 (98) & 72068 (71)  \\
    &  & \textit{16} & 1765 (100)  & 3714 (100) & &  1640 (97) & 4304 (70) \\
    \hline \noalign{\smallskip}
II  & \textbf{2 cluster nodes}  %\multirow{3}{*}{}
  & \textit{1} & 13056 (100) & 22938 (99) & & 10558 (89) & 27128 (73) \\
&   & \textit{16} & 732 (100) & 1759 (100)  &  & 1130 (92) & 2254 (73) \\
    \hline \noalign{\smallskip}
III & \textbf{4 cluster nodes} %\multirow{3}{*}{}
  & \textit{1}  & 5016 (100) & 12703 (100) & & 5777 (94) & 15980 (75) \\
& & \textit{16} & 367 (100) & 776 (100) & & 529 (88) &  1264 (66) \\
    \hline \noalign{\smallskip}
VI & \textbf{8 cluster nodes} %\multirow{3}{*}{}
  & \textit{1}  & 2103 (100) & 5063 (100) & & 2847 (97) & 9853 (83)\\
& & \textit{16} & 145 (100)  & 310 (100)  & & 272 (83) & 774 (57)\\
    \hline \noalign{\smallskip}
V & \textbf{12 cluster nodes} %\multirow{3}{*}{}
  & \textit{1}  & 1155 (100) & 2399 (100) & & 2233 (98) & 6022 (86) \\
& & \textit{16} & 76 (100)  & 159 (100)  & & 168 (87) & 393 (53)\\
    \noalign{\smallskip}\hline
  \end{tabular}
\end{table}

The averaged speedup in the number of iterations and the one in time (in brackets) are presented in
Table \ref{tab:speedup}. In the first row of Table \ref{tab:speedup}, the averaged number of
iterations and the averaged time of solving the problems in the sequential mode (in brackets) are
presented. MIDACO uses a smaller number of communications in the parallel computations.
Therefore, the indicators of speedup increased monotonously with increasing number of
computational nodes employed. In Globalizer, each node should transfer the results of the executed
iterations to all other nodes. Therefore, the efficiency of parallelization decreased with increasing
number of nodes and the number of active threads on each node. This is evident from Table
\ref{tab:speedup}: the speedup in time decreased at the transition from 4 nodes to 8 ones when
using 16 threads. In the case of using 1 thread per a node, the speedup of both optimization methods
were approximately the same. However, it should be noted that Globalizer executed less number of
trials, and spends much less time for solving all problems when \(p=1\). With increasing number of
nodes and threads, MIDACO approaches Globalizer in the averaged time of problem solving,
however, its reliability decreases at the same time.

@@

\begin{table}
  \centering
  \caption{Speedup of parallel computations executed by the parallel algorithm}
  \label{tab:speedup}
  \begin{tabular}{cccccccc}
    \cline{3-8}\noalign{\smallskip}
    \multicolumn{2}{c}{  } & \textit{p} & \multicolumn{2}{c}{Globalizer} & &
\multicolumn{2}{c}{MIDACO}   \\
    \noalign{\smallskip} \cline{4-5} \cline{7-8}  \noalign{\smallskip}
    \multicolumn{2}{c}{  } & & \textit{Simple} & \textit{Hard} & & \textit{Simple} &
\textit{Hard}  \\
    \noalign{\smallskip}\hline
I  & \textbf{1 cluster node}  %\multirow{3}{*}{}
    & \textit{1}   & 25270 (27.3s) & 55180 (61.8s) & & 27645 (32.2s) & 72068 (132.5s)  \\
  &  & \textit{16} & 14.3(11.7) & 14.9(12.6)  & &  16.9(14.4) & 16.7(14.4) \\
  \hline \noalign{\smallskip}
II  & \textbf{2 cluster nodes}  %\multirow{3}{*}{}
  & \textit{1}      &   1.9(1.9) & 2.4(2.2)  & & 2.6(1.7) & 2.7(2.3) \\
  &   & \textit{16} & 34.5(20.4) & 31.4(18.8) & & 24.5(18.8) & 32.0(29.1) \\
    \noalign{\smallskip}\hline	\noalign{\smallskip}
III  & \textbf{4 cluster nodes}  %\multirow{3}{*}{}
& \textit{1}      & 5.0(4.6) & 4.3(4.1) & & 4.8(3.9) & 4.5(4.4) \\
&   & \textit{16} & 68.8(23.7) & 71.2(25.8) & & 52.3(32.9) & 57.0(50.2) \\
  \noalign{\smallskip}\hline	\noalign{\smallskip}
VI & \textbf{8 cluster nodes} %\multirow{3}{*}{}
& \textit{1}    & 12.0(8.7) & 10.9(8.3) & & 9.7(8.2)    & 7.3(9.0)    \\
& & \textit{16} & 174.1(4.7) & 177.8(5.3) & & 101.6(51.6) & 93.1(84.1) \\
    \noalign{\smallskip}\hline
    V & \textbf{12 cluster nodes} %\multirow{3}{*}{}
    & \textit{1}    & 21.9(11.4)  & 23.0(12.8)  & & 12.4(11.1)  & 12.0(14.8)  \\
    & & \textit{16} & 333.5(2.7) & 347.9(3.0) & & 164.5(82.0) & 183.2(129.7) \\
        \noalign{\smallskip}\hline
  \end{tabular}
\end{table}

>>>>

\documentclass{svproc}
%
% RECOMMENDED
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%
%
\usepackage{graphicx}
\usepackage{marvosym}
\usepackage{amssymb}
\usepackage{cite}
\usepackage{subfig}
% to typeset URLs, URIs, and DOIs
\usepackage{url}
\usepackage{hyperref}
\def\UrlFont{\rmfamily}

@@

\begin{abstract}
This work considers several stochastic and deterministic derivative-free global optimization
algorithms. In the first part of the paper popular sequential open-source solvers are compared against
the Globalizer solver, which is developed at the Lobachevsky State University. The Globalizer is
designed to solve problems with black-box objectives satisfying the Lipschitz condition and shows
competitive performance with other similar solvers. The comparison is done on several sets of
challenging multi-extremal benchmark functions. The second part of this work is devoted to
comparison between the Globalizer and MIDACO solvers on systems with shared and distributed
memory. MIDACO is a state-of-the-art global solver included to the TOMLAB optimization
environment for MATLAB. Results of the benchmark show advantages of the Globalizer on small-
dimensional, but sufficiently multi-extremal benchmark functions.
% We would like to encourage you to list your keywords within
% the abstract section using the \keywords{...} command.
\keywords{deterministic global optimization $\cdot$ stochastic global optimization
  $\cdot$ parallel numerical methods $\cdot$ derivative-free algorithms $\cdot$ black-box
optimization}
\end{abstract}
%
\section{Introduction}

@@

Let us consider the averaged number of trials executed to solve a single problem and the number of
solved problems as the characteristics of the optimization method on each class. The less the number
of trials, the faster the method converges to a solution, hence the less times it turns to a potentially
computation-costly procedure of computing the objective function. The number of solved problems
evidences the reliability of the method at given parameters on the class of test problems being
solved. In order to make independent the quantities featuring the reliability and the speed of convergence,
averaged number of trials always was calculated taking into account solved problems only.

@@

Computational experiments with parallel algorithms have been carried out on the Lobachevsky
supercomputer at
State University of Nizhni Novgorod. A computational node includes 2 Intel
Sandy Bridge E5-2660 2.2 GHz processors, 64 GB RAM. Each CPU has 8 cores (i. e. total 16
cores were available per a node). All the computational nodes are working under CentOS Linux 7.2.
All the considered algorithms are implemented using C++ and compiled by GCC 4.8.5.

@@

\begin{table}
  \centering
  \caption{Averaged numbers of iterations executed by the parallel algorithms for solving the test
optimization problems}
  \label{tab:iterations}
  \begin{tabular}{cccccccc}
    \cline{3-8}\noalign{\smallskip}
    \multicolumn{2}{c}{  } & \textit{p} & \multicolumn{2}{c}{Globalizer} & &
\multicolumn{2}{c}{MIDACO}   \\
    \noalign{\smallskip} \cline{4-5} \cline{7-8}  \noalign{\smallskip}
    \multicolumn{2}{c}{  } & & \textit{Simple} & \textit{Hard} & & \textit{Simple} &
\textit{Hard}  \\
    \noalign{\smallskip}\hline
    I & \textbf{1 cluster node}
      & \textit{1} &   25270 (100)  & 55180 (99) & & 27645 (98) & 72068 (71)  \\
    &  & \textit{16} & 1765 (100)  & 3714 (100) & &  1640 (97) & 4304 (70) \\
    \hline \noalign{\smallskip}
II  & \textbf{2 cluster nodes}  %\multirow{3}{*}{}
  & \textit{1} & 13056 (100) & 22938 (99) & & 10558 (89) & 27128 (73) \\
&   & \textit{16} & 732 (100) & 1759 (100)  &  & 1130 (92) & 2254 (73) \\
    \hline \noalign{\smallskip}
III & \textbf{4 cluster nodes} %\multirow{3}{*}{}
  & \textit{1}  & 5016 (100) & 12703 (100) & & 5777 (94) & 15980 (75) \\
& & \textit{16} & 367 (100) & 776 (100) & & 529 (88) &  1264 (66) \\
    \hline \noalign{\smallskip}
VI & \textbf{8 cluster nodes} %\multirow{3}{*}{}
  & \textit{1}  & 2103 (100) & 5063 (100) & & 2847 (97) & 9853 (83)\\
& & \textit{16} & 145 (100)  & 310 (100)  & & 272 (83) & 774 (57)\\
    \hline \noalign{\smallskip}
V & \textbf{12 cluster nodes} %\multirow{3}{*}{}
  & \textit{1}  & 1155 (100) & 2399 (100) & & 2233 (98) & 6022 (86) \\
& & \textit{16} & 76 (100)  & 159 (100)  & & 168 (87) & 393 (53)\\
    \noalign{\smallskip}\hline
  \end{tabular}
\end{table}

The averaged speedup in the number of iterations and the one in time (in brackets) are presented in
Table \ref{tab:speedup}. In the first row of Table \ref{tab:speedup}, the averaged number of
iterations and the averaged time of solving the problems in the sequential mode (in brackets) are
presented. MIDACO uses a smaller number of communications in the parallel computations.
Therefore, the indicators of speedup increased monotonously with increasing number of
computational nodes employed. In Globalizer, each node should transfer the results of the executed
iterations to all other nodes. Therefore, the efficiency of parallelization decreased with increasing
number of nodes and the number of active threads on each node. This is evident from Table
\ref{tab:speedup} and Picture \ref{fig:speedup_p16}: the speedup in time decreased at the transition from 4 nodes to 8 ones when
using 16 threads. In the case of using 1 thread per a node, the speedup of both optimization methods
were approximately the same (see Picture \ref{fig:speedup_p16}). However, it should be noted that Globalizer executed less number of
trials, and spends much less time for solving all problems when \(p=1\). With increasing number of
nodes and threads, MIDACO approaches Globalizer in the averaged time of problem solving (see Pictures \ref{fig:time_p1}, \ref{fig:time_p16}),
however, its reliability decreases at the same time.

@@

\begin{figure}[ht]
  \centering
  \subfloat[$p=1$]{{\includegraphics[width=.5\textwidth]{images/speedup_midaco_globalizer_p_1.pdf}}\label{fig:speedup_p1}}
  \subfloat[$p=16$]{{\includegraphics[width=.5\textwidth]{images/speedup_midaco_globalizer_p_16.pdf}}\label{fig:speedup_p16}}
  \caption{Speedup demonstrated by the parallel algorithms when solving problems from the GKLS 4d Hard class}
\end{figure}

\begin{figure}[ht]
  \centering
  \subfloat[$p=1$]{{\includegraphics[width=.5\textwidth]{images/time_midaco_globalizer_p_1.pdf}}\label{fig:time_p1}}
  \subfloat[$p=16$]{{\includegraphics[width=.5\textwidth]{images/time_midaco_globalizer_p_16.pdf}}\label{fig:time_p16}}
  \caption{Averaged execution time of the parallel algorithms when solving problems from the GKLS 4d Hard class}
\end{figure}

\begin{table}
  \centering
  \caption{Speedup of parallel computations executed by the parallel algorithms}
  \label{tab:speedup}
  \begin{tabular}{cccccccc}
    \cline{3-8}\noalign{\smallskip}
    \multicolumn{2}{c}{  } & \textit{p} & \multicolumn{2}{c}{Globalizer} & &
\multicolumn{2}{c}{MIDACO}   \\
    \noalign{\smallskip} \cline{4-5} \cline{7-8}  \noalign{\smallskip}
    \multicolumn{2}{c}{  } & & \textit{Simple} & \textit{Hard} & & \textit{Simple} &
\textit{Hard}  \\
    \noalign{\smallskip}\hline
I  & \textbf{1 cluster node}  %\multirow{3}{*}{}
    & \textit{1}   & 25270 (27.3s) & 55180 (61.8s) & & 27645 (32.2s) & 72068 (132.5s)  \\
  &  & \textit{16} & 14.3(11.7) & 14.9(12.6)  & &  16.9(14.4) & 16.7(14.4) \\
  \hline \noalign{\smallskip}
II  & \textbf{2 cluster nodes}  %\multirow{3}{*}{}
  & \textit{1}      &   1.9(1.9) & 2.4(2.2)  & & 2.6(1.7) & 2.7(2.3) \\
  &   & \textit{16} & 34.5(20.4) & 31.4(18.8) & & 24.5(18.8) & 32.0(29.1) \\
    \noalign{\smallskip}\hline	\noalign{\smallskip}
III  & \textbf{4 cluster nodes}  %\multirow{3}{*}{}
& \textit{1}      & 5.0(4.6) & 4.3(4.1) & & 4.8(3.9) & 4.5(4.4) \\
&   & \textit{16} & 68.8(23.7) & 71.2(25.8) & & 52.3(32.9) & 57.0(50.2) \\
  \noalign{\smallskip}\hline	\noalign{\smallskip}
VI & \textbf{8 cluster nodes} %\multirow{3}{*}{}
& \textit{1}    & 12.0(8.7) & 10.9(8.3) & & 9.7(8.2)    & 7.3(9.0)    \\
& & \textit{16} & 174.1(4.7) & 177.8(5.3) & & 101.6(51.6) & 93.1(84.1) \\
    \noalign{\smallskip}\hline
    V & \textbf{12 cluster nodes} %\multirow{3}{*}{}
    & \textit{1}    & 21.9(11.4)  & 23.0(12.8)  & & 12.4(11.1)  & 12.0(14.8)  \\
    & & \textit{16} & 333.5(2.7) & 347.9(3.0) & & 164.5(82.0) & 183.2(129.7) \\
        \noalign{\smallskip}\hline
  \end{tabular}
\end{table}
